---
title: "GAE7"
author: "Wyatt Stephenson"
format: 
  html:
    theme: custom.scss
    embed-resources: true
    code-tools: true

csl: ecology.csl

editor: visual
bibliography: references.bib
---

```{r}
#| label: SetUp
#| message: FALSE
#| echo: FALSE
#| warning: false

knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE
  ) 

library(emmeans)
library(lmerTest)
library(effectsize)
library(car)
library(ggfortify)
library(patchwork)
library(broom)
library(latex2exp)       
library(kableExtra)
library(tidyverse)

options(knitr.kable.NA = '')
```

## Introduction

@Pintar.etal2018 conducted a field mesocosm experiment to explore the relative effects of predation risk and nutrient availability (a pdf of their publication is included in the GitHub repository). The predation risk manipulation was accomplished by adding either 0, 1, or 2, Golden Topminnows (*Fundulus chrysotus*) to each mesocosm. These small fish are surface-feeding predators and are known to elicit colonization and oviposition avoidance among dytiscid beetles and *Culex* mosquitoes, but not in some hydrophilid beetles. Nutrient-level treatments were created by the addition of 0, 4, or 8 g of rabbit chow (Small World Rabbit Food, Mannapro, St. Louis, MO, 40% protein).

## Methods

Please see @Pintar.etal2018 for greater detail on experimental design and data collection. In brief, the experiment was set up as a 3 X 3 factorial treatment design (9 treatment combinations). The experiment was replicated in 3 separate locations at 3 different times for a total of 9 replicates. This should have resulted in 81 observations total, however some fish died during the experiment rendering those observations not being included in the data resulting in only 69 observations.

```{r }
#| label: dataImport

df1 <- read_csv("Pintar_etal_2018.csv", 
                show_col_types = FALSE)

# Block = 1 to 9 for the 9 spatiotemporal blocks (3 locations X 3 times)
# Pool = 1 to 9 for the plastic wading pools used for the experiment
# Location = 1 to 3 for the 3 fields where the pools were deployed
# Time = a to c for the 3 time periods used for each 2-week run of the exp.
# Nutrient = 1 to 3 for the 3 nutrient levels
# Fish = 1 to 3 for the 3 fish densities
# Culex_Egg_Rafts = number of egg rafts collected over the 2 weeks
# Hydrophilidae = number of hydrophilid beetles (Water Scavenger Beetles)
# Dytiscidae = number of dytiscid beetles (Predaceous Diving Beetles)

# Fish and Nutrient are provided as integers, but I wish to use them as ordered
# factors so that, by default, polynomial contrasts will be used to compute the
# ANOVAs.

df1$Nutrient <- factor(df1$Nutrient)

df1$Fish <- factor(df1$Fish
                   )

df1$Block <- factor(df1$Block)

df1$Pool <- factor(df1$Pool)

df1$Location <- factor(df1$Location)

df1$Time <- factor(df1$Time)

```

### Statistical Methods

First three linear models were created comparing Culex egg rafts, Hydrophilidae, and Dytiscidae against predation and nutrient levels along with the interaction between predation and nutrient levels using the @lmer package. Next a standard ANOVA was run for all three liner models using base @R . Lastly both a fixed effect and random effect ANOVA was run for all three linear models using the @lmer package.

## Results

When looking at the results of our fixed affect ANOVAs we can see that only predation level had any significance and this was only in realtion to amount of Culex egg rafts and amount of Dytiscidae @tbl-FixedAnova and @tbl-FixedAnova3 . When comparing these fixed affect modelss to a random affect ANOVA we can see that the the fixed affect was the stronger model as seen by all three of the fixed affect modelss having a higher AIC score than the random affect ( @tbl-MixedAnova, @tbl-MixedAnova2, @tbl-MixedAnova3) and none of the AIC scores being even close to two unit delta need to consider them as better models,

```{r}
#| label: tbl-dataImport
#| tbl-cap: "First six observations of data provided by @Pintar.etal2018"

colnames(df1) <- tolower(colnames(df1))


kbl(
  head(df1),
  col.names = c(
    "Block",
    "Pool",
    "Location",
    "Time",
    "Nutrient",
    "Fish",
    "Culex Egg Rafts",
    "Hydrophilidae",
    "Dytisicidae"
  )
) |>
  kable_classic(
    full = FALSE
  )
```

```{r}
#| label: tbl-Culex
#| tbl-cap: "Effects of predation and nutrient levels on Culfex egg rafts"
#| fig-show: hide
#| results: hide

CulexFactorial <- lmer(culex_egg_rafts ~ fish*nutrient + (1 | block), data = df1)

CulexAnova <- (anova(CulexFactorial))

kbl(
  CulexAnova, col.names = c(
    "SS",
    "MS",
    "df1",
    "df2",
    "F",
    "p"
    
  ),
  digits = c(3,3,0,0,3,4)
) |>
  kable_classic()


DytisFactorial <- lmer(dytiscidae ~ fish*nutrient + (1 | block), data = df1)

DytisAnova <- (anova(DytisFactorial))

kbl(
  DytisAnova, col.names = c(
    "SS",
    "MS",
    "df1",
    "df2",
    "F",
    "p"
    
  ),
  digits = c(3,3,0,0,3,4)
) |>
  kable_classic()

HydroFactorial <- lmer(hydrophilidae ~ fish*nutrient + (1 | block), data = df1)

HydroAnova <- (anova(HydroFactorial))

kbl(
  HydroAnova, col.names = c(
    "SS",
    "MS",
    "df1",
    "df2",
    "F",
    "p"
    
  ),
  digits = c(3,3,0,0,3,4)
) |>
  kable_classic()


```

```{r}
#| label: tbl-FixedAnova
#| tbl-cap: " Fixed ANOVA summary for linear model of Culex egg rafts aginst predation and nutirent levels "

kbl(
  anova(CulexFactorial),
  digits = c(3,0,3,3,4)
) |> kable_classic()

```

```{r}
#| label: tbl-FixedAnova2
#| tbl-cap: " Fixed ANOVA summary for linear model of Hydrophilidae aginst predation and nutirent levels "

kbl(
  anova(HydroFactorial),
  digits = c(3,0,3,3,4)
) |> kable_classic()
```

```{r}
#| label: tbl-FixedAnova3
#| tbl-cap: " Fixed ANOVA summary for linear model of Dytiscidae aginst predation and nutirent levels "

kbl(
  anova(DytisFactorial),
  digits = c(3,0,3,3,4)
) |> kable_classic()
```

```{r}
#| label: tbl-MixedAnova
#| tbl-cap: "Linear mixed-effect model results for Culex egg rafts and predation and nutrient levels"
#| tbl-subcap:
#|     - "Fixed effect"
#|     - "Random effect"
#| layout-nrow: 2

CulexMixed <- anova(CulexFactorial)

kbl(
  CulexMixed,
  col.names = c(
    "SS",
    "MS",
    "df1",
    "df2",
    "F",
    "p"
  ),
  digits = c(3,3,0,0,3,4)
) |>
  kable_classic()

kbl(
  rand(CulexFactorial),
  col.names = c(
    "Parameters",
    "log-liklehood",
    "AIC",
    "LRT",
    "df",
    "p"
  ),
  digits = c(0,3,3,3,0,4)
)|>
  kable_classic()

```

```{r}
#| label: tbl-MixedAnova2
#| tbl-cap: "Linear mixed-effect model results for Hydrophilidae and predation and nutrient levels"
#| tbl-subcap:
#|     - "Fixed effect"
#|     - "Random effect"
#| layout-nrow: 2

HydroMixed <- anova(HydroFactorial)

kbl(
  HydroMixed,
  col.names = c(
    "SS",
    "MS",
    "df1",
    "df2",
    "F",
    "p"
  ),
  digits = c(3,3,0,0,3,4)
) |>
  kable_classic()

kbl(
  rand(HydroFactorial),
  col.names = c(
    "Parameters",
    "log-liklehood",
    "AIC",
    "LRT",
    "df",
    "p"
  ),
  digits = c(0,3,3,3,0,4)
)|>
  kable_classic()
```

```{r}
#| label: tbl-MixedAnova3
#| tbl-cap: "Linear mixed-effect model results for Dytiscidae and predation and nutrient levels"
#| tbl-subcap:
#|     - "Fixed effect"
#|     - "Random effect"
#| layout-nrow: 2

DytisMixed <- anova(DytisFactorial)

kbl(
  DytisMixed,
  col.names = c(
    "SS",
    "MS",
    "df1",
    "df2",
    "F",
    "p"
  ),
  digits = c(3,3,0,0,3,4)
) |>
  kable_classic()

kbl(
  rand(DytisFactorial),
  col.names = c(
    "Parameters",
    "log-liklehood",
    "AIC",
    "LRT",
    "df",
    "p"
  ),
  digits = c(0,3,3,3,0,4)
)|>
  kable_classic()
```

## References
